---
title: "An interactive Formula 1 Analysis"
author: "Nomthunzi Moyo and Mateusz Majak"
date: "22 11 2020"
output: 
  html_document:
    number_sections: true
    fig_caption: true
    toc: true
    fig_width: 7
    fig_height: 4.5
    theme: cosmo
    highlight: tango
    code_folding: hide
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# About the dataset
The dataset used in this project consists of information about F1 races statistics
from 1950 until the latest 2020 season. It was shared by Vopani on the Kaggle 
platform and is available [here](https://www.kaggle.com/rohanrao/formula-1-world-championship-1950-2020).
The whole data comes from [ergast website](http://ergast.com/mrd/) and is divided into
13 csv files:

* circuits
* constructor_results
* constructor_standings
* constructors
* driver_standings
* drivers
* lap_times
* pit_stops
* qualifying
* races
* results
* seasons
* status

It is a real challenge to compare drivers, cars and teams, considering very complex structure of the car and the number of changing factors on the track.

With such amount of data, we will try to answer the following questions:

* **How has the competition and level of F1 changed over the years?**
* **Is F1 currently being won in qualifying?**
* **Do the differences between the cars decrease over the years?**
* **Which F1 circuits provide a more even fight?**

# Preparations {.tabset .tabset-fade .tabset-pills}

## Load libraries
We load multiple libraries for data manipulation and visualisation together with 
tools for data modelling.

```{r message = FALSE}

# data manipulation
library(dplyr)
library(tidyr)

# Interactivity
library(plotly)
library(crosstalk)
library(DT)

#animations
library(gganimate)

#Shiny App
library(shiny)

#visualisations
library(ggplot2)
library(RColorBrewer)
library(ggthemes)
library(scales)
library(extrafont)
loadfonts()

#maps
library(ggmap)
library(maps)

```

## Load Data
```{r message = FALSE}
#results - main csv file
results <- read.csv('F1_data/results.csv',sep=',', stringsAsFactors=F)

#constructor_results
constructor_results <- read.csv('F1_data/constructor_results.csv',sep=',', stringsAsFactors=F)

#constructor_standings
constructor_standings <- read.csv('F1_data/constructor_standings.csv',sep=',', stringsAsFactors=F)

#constructors
constructors <- read.csv('F1_data/constructors.csv',sep=',', stringsAsFactors=F)

#driver_standings
driver_standings <- read.csv('F1_data/driver_standings.csv',sep=',', stringsAsFactors=F)

#drivers
drivers <- read.csv('F1_data/drivers.csv',sep=',', stringsAsFactors=F)

#lap_times
lap_times <- read.csv('F1_data/lap_times.csv',sep=',', stringsAsFactors=F)

#pit_stops
pit_stops <- read.csv('F1_data/pit_stops.csv',sep=',', stringsAsFactors=F)

#qualifying
qualifying <- read.csv('F1_data/qualifying.csv',sep=',', stringsAsFactors=F)

#races
races <- read.csv('F1_data/races.csv',sep=',', stringsAsFactors=F)

#seasons
seasons <- read.csv('F1_data/seasons.csv',sep=',', stringsAsFactors=F)

#status
status <- read.csv('F1_data/status.csv',sep=',', stringsAsFactors=F)

#circuits
circuits <- read.csv('F1_data/circuits.csv',sep=',', stringsAsFactors=F)
```

## Data manipulation
```{r}
#results - main csv file
results$fastestLapSpeed[which(results$fastestLapSpeed=="\\N")] <- NA
results$fastestLapSpeed <- as.numeric(results$fastestLapSpeed)

#constructor_results_merged
constructor_results_merged <- left_join(
  constructor_results, constructors %>%
    dplyr::select(-url) %>%
    dplyr::rename(constructor = name),
  by = 'constructorId'
)
constructor_results_merged <- left_join(
  constructor_results_merged, races %>%
    dplyr::rename(race_name = name),
  by = 'raceId'
)
constructor_results_merged <- left_join(
  constructor_results_merged, constructor_standings %>%
    dplyr::rename(constructor_points = points),
  by = c('constructorId','raceId')
)
constructor_results_merged$date <- as.Date(constructor_results_merged$date)

#drivers
drivers <- left_join(
  drivers %>%
    dplyr::select(-url), driver_standings,
  by = 'driverId')

#races
races$date<-as.Date(races$date)
races$name<-gsub("Grand Prix", "GP", races$name)

#results_races
races <- left_join(
  races %>%
    dplyr::select(-name, -url), 
  circuits %>%
    dplyr::select(-url),
  by = 'circuitId')

results_races <- left_join(
  results %>%
    dplyr::select(-time), 
  races %>%
    dplyr::select(-time),
  by = 'raceId')

#results_drivers
results_drivers <- left_join(
    results, 
    drivers %>%
      dplyr::select(-points, -position, -positionText),
    by = c('driverId', 'raceId')) 
    
results_drivers <- left_join(
  results_drivers, races %>%
    dplyr::select(-time),
  by = 'raceId')
```

# Data preview {.tabset .tabset-fade .tabset-pills}

## Results
```{r}
datatable(results[1:100,] %>%                      
    arrange(resultId), 
    options = list(pageLength = 10, scrollX = T)) 
```

## constructor_results
```{r}
datatable(constructor_results[1:100,] %>%                      
    arrange(constructorResultsId), 
    options = list(pageLength = 10, scrollX = T)) 
```

## constructor_standings
```{r}
datatable(constructor_standings[1:100,] %>%                      
    arrange(constructorStandingsId), 
    options = list(pageLength = 10, scrollX = T)) 
```

## constructors
```{r}
datatable(constructors[1:100,] %>%                      
    arrange(constructorId), 
    options = list(pageLength = 10, scrollX = T)) 
```

## driver_standings
```{r}
datatable(driver_standings[1:100,] %>%                      
    arrange(driverStandingsId), 
    options = list(pageLength = 10, scrollX = T)) 
```

## drivers
```{r}
datatable(drivers[1:100,] %>%                      
    arrange(driverId), 
    options = list(pageLength = 10, scrollX = T)) 
```

## lap_times
```{r}
datatable(lap_times[1:100,] %>%                      
    arrange(raceId), 
    options = list(pageLength = 10, scrollX = T)) 
```

## pit_stops
```{r}
datatable(pit_stops[1:100,] %>%                      
    arrange(raceId), 
    options = list(pageLength = 10, scrollX = T)) 
```

## qualifying
```{r}
datatable(qualifying[1:100,] %>%                      
    arrange(qualifyId), 
    options = list(pageLength = 10, scrollX = T)) 
```

## races
```{r}
datatable(races[1:100,] %>%                      
    arrange(raceId), 
    options = list(pageLength = 10, scrollX = T)) 
```

## seasons
```{r}
datatable(seasons[1:100,] %>%                      
    arrange(year), 
    options = list(pageLength = 10, scrollX = T)) 
```

## status
First 100 rows
```{r}
datatable(status[1:100,] %>%                      
    arrange(statusId), 
    options = list(pageLength = 10, scrollX = T)) 
```

# EDA

## F1 across the World

In the last 70 years, the competition in F1 has been held on 6 continents.
```{r, warning=FALSE, fig.width=10}
#circuits map
#Change longitude column name for cohesion
colnames(results_races)[27] <- "lon"
# generate worldwide dots
lat <- data_frame(lat = seq(-90, 90, by = 1.5))
lon <- data_frame(lon = seq(-170, 180, by = 1.5))
dots <- merge(lat, lon, all = TRUE)  

# exclude water-based dots 
dots %>% 
  mutate(country = map.where("world", lon, lat),
         lakes = map.where("lakes", lon, lat)) %>%
  filter(!is.na(country) & is.na(lakes)) %>% 
  select(-lakes) ->
  dots

# plot the data
ggplot(dots, aes(x = lon, y = lat)) + 
geom_point(col = "white", size = 0.25) +
geom_point(data = results_races, colour = "red", alpha = .009, size = 5) + 
theme_void() +
theme(panel.background = element_rect(fill = "#484848"),
      plot.background = element_rect(fill = "#484848"))
```


## Question A - How has the competition and level of F1 changed over the years?

### Running sum of points for top 10 constructors
```{r, warning=FALSE, message=FALSE, fig.width=10}
# Teams running sum by wins/championships
top10teams <- (aggregate(points ~ constructor, constructor_results_merged, sum) %>% 
  arrange(desc(points)) %>% 
  top_n(10))$constructor

constructor_results_merged$top10 <- ifelse(
  constructor_results_merged$constructor %in% top10teams,
  constructor_results_merged$constructor,'Other')

# Static plot
# constructor_results_merged %>%
#   dplyr::filter(top10 != 'Other') %>%
#   dplyr::select(year, date, constructor, points) %>%
#   dplyr::group_by(constructor) %>% 
#   dplyr::arrange(date) %>% 
#   dplyr::mutate(point_sum = cumsum(points)) %>%
#   dplyr::group_by(year, constructor) %>%
#   dplyr::summarise(point_sum=max(point_sum)) %>% 
#   ggplot(aes(x = year, y = point_sum, col = constructor, group = constructor)) +
#   geom_line(size=1, alpha = .5) +
#   geom_point(alpha=.5,size=.2)

constructor_results_plotly <- highlight_key(constructor_results_merged %>% 
  dplyr::filter(top10 != 'Other') %>% 
  dplyr::select(year, date, constructor, points) %>% 
  dplyr::group_by(constructor) %>% 
  dplyr::arrange(date) %>% 
  dplyr::mutate(PointSum = cumsum(points)) %>% 
  dplyr::group_by(year, constructor) %>% 
  dplyr::summarise(PointSum=max(PointSum)))

gg <- constructor_results_plotly %>% 
  ggplot(aes(year, PointSum, col = constructor, group = constructor)) +
  geom_line(size=1, alpha = .5) +
  geom_point(alpha=.5,size=.2)

labs(x = "Year", y = "constructor") +
  theme_tufte() +
  NULL

filter <- bscols(filter_select("constructor","Sum of points over years (select a constructor):",
                               constructor_results_plotly, ~constructor, 
                               multiple = TRUE),
                 ggplotly(gg, dynamicTicks = TRUE),
                 widths = c(15, 15)
)
bscols(filter)
```

```{r, fig.width=10}
#constructors running sum animated plot

# Top teams running sum by wins/championships
top10teams <- (aggregate(points ~ constructor, constructor_results_merged, sum) %>% 
                 arrange(desc(points)) %>% 
                 top_n(10))$constructor

constructor_results_merged$top10 <- ifelse(
  constructor_results_merged$constructor %in% top10teams,
  constructor_results_merged$constructor,'Other')

team_colors <- c("Benetton" = "#21783b", "Ferrari" = "#cf0e0e", "Force India" = "#d81676", 
                 "Lotus F1" = "#010101", 'McLaren' = '#ffa113', 'Mercedes' = '#1fb1a7',
                 'Red Bull' = '#00007e', 'Renault' = '#ffd700', 'Team Lotus' = '#b28c14',
                 "Williams" = '#ffffff')

anim <- constructor_results_merged %>%
  dplyr::filter(top10 != 'Other') %>%
  dplyr::select(year, date, constructor, points) %>%
  dplyr::group_by(constructor) %>% 
  dplyr::arrange(date) %>% 
  dplyr::mutate(point_sum = cumsum(points)) %>%
  dplyr::group_by(year, constructor) %>%
  dplyr::summarise(point_sum=max(point_sum)) %>% 
  ggplot(aes(x = year, y = point_sum, col = constructor, group = constructor)) +
  geom_step(size=1, alpha = 1) +
  geom_point(alpha=.5,size=.2) + 
  scale_color_manual(values = team_colors) +
  theme(plot.background = element_rect(fill = 'lightgrey'),
        axis.line = element_line(color = 'black'),
        panel.grid = element_line(color = 'grey'),
        plot.title = element_text(size = 46, hjust = .5, face = 'bold', color = '#cf0e0e'),
        axis.title = element_text(size = 19),
        legend.title = element_text(size = 19),
        legend.text = element_text(size = 17),
        legend.background = element_rect(fill = 'lightgrey'),
        legend.key = element_rect(color = 'grey')) +
  labs(x = 'Year',
       y = 'Points sum',
       title = 'Running sum of constructors\' points') +
  guides(color = guide_legend(title = 'Constructor')) + # now i add frame variable
  gganimate::transition_reveal(year)

animate(anim, height = 600, width =1080, nframes = 160, fps = 5)

```

### Top 25 drivers
```{r warning=FALSE, message=FALSE, fig.width=10}
results_drivers %>% 
  dplyr::select(points, driverRef, forename, surname, nationality, year) %>% 
  tidyr::unite(full_name, c('forename', 'surname'), sep = ' ') %>% 
  dplyr::group_by(full_name) %>% 
  dplyr::summarise(year_points = sum(points)) %>% 
  dplyr::filter(year_points > 0) %>% 
  dplyr::top_n(25, year_points) %>% 
  dplyr::arrange(year_points) %>% 
  dplyr::mutate(full_name=factor(full_name, full_name)) %>%
  ggplot(aes(x=full_name, y=year_points)) +
  geom_segment( aes(x=full_name, xend=full_name, y=0, yend=year_points), color = 'grey') +
  geom_point(color = 'red', size = 3) +
  coord_flip() +
  labs(x = 'Driver',
       y = 'Sum of points',
       title = 'Top 25 drivers by number of points in the career') +
  theme(plot.title = element_text(hjust = .5))
```
## Question B: Is F1 currently being won in qualifying?
```{r warning=FALSE, message=FALSE, fig.width=10}
quali_races <- left_join(
  results_races %>%
    dplyr::mutate(race_position = positionOrder,
                  grid_position = grid) %>% 
    dplyr::select(driverId, raceId, date, grid_position, race_position),
  qualifying %>%
    dplyr::mutate(quali_position = position) %>% 
    dplyr::select(driverId, raceId, quali_position),
  by = c('raceId','driverId'))
#101 points to gain from 1st to 10th place
# quali_races %>% 
#   dplyr::select(date, quali_position, race_position)

quali_races %>% 
  dplyr::select(date, grid_position, race_position) %>%
  distinct() %>%
  dplyr::mutate(
    race_points = case_when(race_position == 1  ~ 25,
                            race_position == 2  ~ 18,
                            race_position == 3  ~ 15,
                            race_position == 4  ~ 12,
                            race_position == 5  ~ 10,
                            race_position == 6  ~ 8,
                            race_position == 7  ~ 6,
                            race_position == 8  ~ 4,
                            race_position == 9  ~ 2,
                            race_position == 10 ~ 1,
                            TRUE                ~ 0
                            ),
    grid_points = case_when(grid_position == 1  ~ 25,
                            grid_position == 2  ~ 18,
                            grid_position == 3  ~ 15,
                            grid_position == 4  ~ 12,
                            grid_position == 5  ~ 10,
                            grid_position == 6  ~ 8,
                            grid_position == 7  ~ 6,
                            grid_position == 8  ~ 4,
                            grid_position == 9  ~ 2,
                            grid_position == 10 ~ 1,
                            TRUE                ~ 0
                            )
    ) %>% 
  dplyr::mutate(point_difference = race_points - grid_points) %>% 
  dplyr::select(date, point_difference) %>% 
  dplyr::filter(point_difference > 0) %>% 
  dplyr::group_by(date) %>% 
  dplyr::summarise(won_points_percentage = sum(point_difference)/101) %>% 
  dplyr::arrange(date) %>%
  ggplot(aes(x = date, y = won_points_percentage)) +
  geom_line(size = .5, alpha = .5,  color = 'grey') +
  geom_point(size = .5, alpha = .9, color = 'grey') +
  geom_smooth(se = TRUE,
              color = 'red') +
  labs(x = 'Year',
       y = '% of points gain after qualifying',
       title = 'Percent of additional points gained after qualifying') +
  scale_x_date(name = paste('Years', substr(min(quali_races$date),1,4),'-',substr(max(quali_races$date),1,4)),
               date_breaks = "10 years", date_labels = "%Y",
               date_minor_breaks = "5 years",
               limits = c(min(quali_races$date), max = max(quali_races$date)),
               expand=c(0,120)) +
  scale_y_continuous(labels=scales::percent) +
  theme(plot.title = element_text(hjust = .5))


```

## Rounds that took the most time to complete (worst 10)
# This plot is a scatter plot that illustrates the longest time it tool to complete a round in years
```{r}
# Grouped Scatter Plot
# xoxo<- races %>% group_by(round)
# xoxo
# xoxo2<-head(xoxo[order(xoxo$time, decreasing= T),], n = 10)
# xoxo2
# ggplot(xoxo2, aes(x=round, y=year, color=year)) + 
#     geom_point(size=6)+ scale_fill_gradientn(colours=rainbow(100))
```

## Question D - Which F1 circuits provide a more even fight?

#Top 3 teams comparison

To answer the given question, we plot the average results for each team from the top.
It is highly possible that the circuits where the differences are lower provide more
even fight.
```{r, fig.width=10}
#Data preparation

#lap times with constructors data joining
# top_times <- left_join(
#   constructor_results_merged %>%
#     dplyr::select(raceId, constructor, race_name, date),
#   lap_times %>%
#     dplyr::select(driverId, raceId, lap, milliseconds),
#   by = c('raceId'))

top_times <- 
  left_join(
    left_join(
      left_join(
        lap_times %>%
          dplyr::select(-position, -time),
        results %>%
          dplyr::select(resultId, raceId, driverId, constructorId),
        by = c('raceId','driverId')),
      constructors %>% 
        dplyr::select(-nationality, -url) %>% 
        dplyr::rename(constructor_name = name),
      by = 'constructorId'),
    races %>% 
      dplyr::select(raceId, name, year) %>% 
      dplyr::rename(race_name = name),
    by = 'raceId')

top_times <- top_times %>% 
  dplyr::select(-driverId, -lap, -raceId) %>% 
  dplyr::filter(milliseconds > 0) %>% 
  dplyr::group_by(race_name, constructorId) %>% 
  dplyr::filter(milliseconds > quantile(milliseconds, 0.05)) %>% 
  dplyr::filter(milliseconds < quantile(milliseconds, 0.95)) %>% 
  dplyr::mutate(scaled_time = (milliseconds-min(milliseconds))/(max(milliseconds)-min(milliseconds))) %>% 
  dplyr::group_by(race_name, constructorId) %>% 
  dplyr::mutate(median_time = median(scaled_time))

# Plot colors
team_colors <- c("Ferrari" = "#cf0e0e", 'Mercedes' = '#1fb1a7',
                 'Red Bull' = '#00007e', 'McLaren' = '#ff8700')

times_theme <- 
    theme(
      legend.position="none",
      panel.background = element_rect(fill = "#3a3c3d"),
      panel.grid = element_line(color = '#2e2f30'),
      panel.grid.major  = element_line(color = '#3a3c3d'),
      plot.background = element_rect(fill = '#3a3c3d'),
      axis.text = element_text(color = 'white'),
      axis.title = element_text(color = 'white'),
      plot.title = element_text(size = 20, color = 'white', hjust = .5)
    ) 

```

``` {r, fig.width=10, fig.height=20}
#Plots 2019
top_times %>% 
  dplyr::filter(constructor_name %in% c('Red Bull','Ferrari','Mercedes')) %>%
  dplyr::filter(year == 2019) %>% 
  ggplot(aes(x=scaled_time, color=constructor_name)) +
  geom_density(alpha=0.6, size = 1) +
  geom_vline(aes(xintercept = median_time, color = constructor_name),linetype="dashed",
             size = 1) +
  scale_color_manual(values = team_colors) +
  times_theme +
  labs(title = '2019 season results for top 3 teams') +
  ylab("Number of laps") +
  xlab("Scaled time") +
  facet_wrap(~race_name, ncol = 3)

# unique((top_times %>% 
#          dplyr::filter(year == 2019))[[6]])

#We can see 4 races with most similar times
similar_results_2019 <- c('Marina Bay Street Circuit','Circuit de Spa-Francorchamps',
                          'Circuit Gilles Villeneuve','Circuit Paul Ricard')
```

```{r fig.width=10}
#Plots 2019 - similar results
top_times %>% 
  dplyr::filter(constructor_name %in% c('Red Bull','Ferrari','Mercedes')) %>%
  dplyr::filter(year == 2019, race_name %in% similar_results_2019) %>% 
  ggplot(aes(x=scaled_time, color=constructor_name)) +
  geom_density(alpha=0.6, size = 1) +
  geom_vline(aes(xintercept = median_time, color = constructor_name),linetype="dashed",
             size = 1) +
  scale_color_manual(values = team_colors) +
  times_theme +
  labs(title = 'Top 4 of most even races in 2019 season') +
  ylab("Number of laps") +
  xlab("Scaled time") +
  facet_wrap(~race_name, ncol = 2)
```

``` {r, fig.width=10, fig.height=20}
#Plots 2020
top_times %>% 
  dplyr::filter(constructor_name %in% c('Red Bull','McLaren','Mercedes')) %>%
  dplyr::filter(year == 2020) %>% 
  ggplot(aes(x=scaled_time, color=constructor_name)) +
  geom_density(alpha=0.6, size = 1) +
  geom_vline(aes(xintercept = median_time, color = constructor_name),linetype="dashed",
             size = 1) +
  scale_color_manual(values = team_colors) +
  times_theme +
  labs(title = '2020 season results for top 3 teams') +
  ylab("Number of laps") +
  xlab("Scaled time") +
  facet_wrap(~race_name, ncol = 3)

# unique((top_times %>% 
#          dplyr::filter(year == 2020))[[6]])

#We can see 4 races with most similar times
similar_results_2020 <- c('Autodromo Internazionale del Mugello',
                     'Circuit de Spa-Francorchamps', 'Red Bull Ring','Sochi Autodrom')
```

```{r fig.width=10}
#Plots 2020 - similar results
top_times %>% 
  dplyr::filter(constructor_name %in% c('Red Bull','McLaren','Mercedes')) %>%
  dplyr::filter(year == 2020, race_name %in% similar_results_2020) %>% 
  ggplot(aes(x=scaled_time, color=constructor_name)) +
  geom_density(alpha=0.6, size = 1) +
  geom_vline(aes(xintercept = median_time, color = constructor_name),linetype="dashed",
             size = 1) +
  scale_color_manual(values = team_colors) +
  times_theme +
  labs(title = 'Top 4 of most even races in 2020 season') +
  ylab("Number of laps") +
  xlab("Scaled time") +
  facet_wrap(~race_name, ncol = 2)
```
```





