---
title: "An interactive Formula 1 Analysis"
author: "Nomthunzi Moyo and Mateusz Majak"
date: "22 11 2020"
output: 
  html_document:
    number_sections: true
    fig_caption: true
    toc: true
    fig_width: 7
    fig_height: 4.5
    theme: cosmo
    highlight: tango
    code_folding: hide
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# About the dataset
The dataset used in this project consists of information about F1 races statistics
from 1950 until the latest 2020 season. It was shared by Vopani on the Kaggle 
platform and is available [here](https://www.kaggle.com/rohanrao/formula-1-world-championship-1950-2020).
The whole data comes from [ergast website](http://ergast.com/mrd/) and is divided into
13 csv files:

* circuits
* constructor_results
* constructor_standings
* constructors
* driver_standings
* drivers
* lap_times
* pit_stops
* qualifying
* races
* results
* seasons
* status

It is a real challenge to compare drivers, cars and teams, considering very complex structure of the car and the number of changing factors on the track.

With such amount of data, we will try to answer the following questions:

* **How has the competition and level of F1 changed over the years?**
* **Is F1 currently being won in qualifying?**
* **Do the differences between the cars decrease over the years?**
* **Which F1 circuits provide a more even fight?**

# Preparations {.tabset .tabset-fade .tabset-pills}

## Load libraries
We load multiple libraries for data manipulation and visualisation together with 
tools for data modelling.

```{r message = FALSE}

# data manipulation
library(dplyr)
library(tidyr)

# Interactivity
library(plotly)
library(crosstalk)
library(DT)

#Shiny App
library(shiny)

#visualisations
library(ggplot2)
library(RColorBrewer)
library(ggthemes)

#maps
library(ggmap)
library(maps)

```

## Load Data
```{r message = FALSE}
#results - main csv file
results <- read.csv('F1_data/results.csv',sep=',', stringsAsFactors=F)

#constructor_results
constructor_results <- read.csv('F1_data/constructor_results.csv',sep=',', stringsAsFactors=F)

#constructor_standings
constructor_standings <- read.csv('F1_data/constructor_standings.csv',sep=',', stringsAsFactors=F)

#constructors
constructors <- read.csv('F1_data/constructors.csv',sep=',', stringsAsFactors=F)

#driver_standings
driver_standings <- read.csv('F1_data/driver_standings.csv',sep=',', stringsAsFactors=F)

#drivers
drivers <- read.csv('F1_data/drivers.csv',sep=',', stringsAsFactors=F)

#lap_times
lap_times <- read.csv('F1_data/lap_times.csv',sep=',', stringsAsFactors=F)

#pit_stops
pit_stops <- read.csv('F1_data/pit_stops.csv',sep=',', stringsAsFactors=F)

#qualifying
qualifying <- read.csv('F1_data/qualifying.csv',sep=',', stringsAsFactors=F)

#races
races <- read.csv('F1_data/races.csv',sep=',', stringsAsFactors=F)

#seasons
seasons <- read.csv('F1_data/seasons.csv',sep=',', stringsAsFactors=F)

#status
status <- read.csv('F1_data/status.csv',sep=',', stringsAsFactors=F)

#circuits
circuits <- read.csv('F1_data/circuits.csv',sep=',', stringsAsFactors=F)
```

## Data manipulation
```{r}
#results - main csv file
results$fastestLapSpeed[which(results$fastestLapSpeed=="\\N")] <- NA
results$fastestLapSpeed <- as.numeric(results$fastestLapSpeed)

#constructor_results


#constructor_standings

#constructors

#driver_standings

#drivers

#lap_times

#pit_stops

#qualifying

#races
races$date<-as.Date(races$date)
races$name<-gsub("Grand Prix","GP",races$name)

#seasons

#status

#connecting main sets
races <- left_join(
  races %>%
    select(-name, -url), 
  circuits %>%
    select(-url), by='circuitId')

results_races<-left_join(
  results %>%
    dplyr::select(-time), 
  races %>%
    dplyr::select(-time), by='raceId')

```

# Data preview {.tabset .tabset-fade .tabset-pills}

## Results
```{r}
datatable(results[1:100,] %>%                      
    arrange(resultId), 
    options = list(pageLength = 10, scrollX = T)) 
```

## constructor_results
```{r}
datatable(constructor_results[1:100,] %>%                      
    arrange(constructorResultsId), 
    options = list(pageLength = 10, scrollX = T)) 
```

## constructor_standings
```{r}
datatable(constructor_standings[1:100,] %>%                      
    arrange(constructorStandingsId), 
    options = list(pageLength = 10, scrollX = T)) 
```

## constructors
```{r}
datatable(constructors[1:100,] %>%                      
    arrange(constructorId), 
    options = list(pageLength = 10, scrollX = T)) 
```

## driver_standings
```{r}
datatable(driver_standings[1:100,] %>%                      
    arrange(driverStandingsId), 
    options = list(pageLength = 10, scrollX = T)) 
```

## drivers
```{r}
datatable(drivers[1:100,] %>%                      
    arrange(driverId), 
    options = list(pageLength = 10, scrollX = T)) 
```

## lap_times
```{r}
datatable(lap_times[1:100,] %>%                      
    arrange(raceId), 
    options = list(pageLength = 10, scrollX = T)) 
```

## pit_stops
```{r}
datatable(pit_stops[1:100,] %>%                      
    arrange(raceId), 
    options = list(pageLength = 10, scrollX = T)) 
```

## qualifying
```{r}
datatable(qualifying[1:100,] %>%                      
    arrange(qualifyId), 
    options = list(pageLength = 10, scrollX = T)) 
```

## races
```{r}
datatable(races[1:100,] %>%                      
    arrange(raceId), 
    options = list(pageLength = 10, scrollX = T)) 
```

## seasons
```{r}
datatable(seasons[1:100,] %>%                      
    arrange(year), 
    options = list(pageLength = 10, scrollX = T)) 
```

## status
First 100 rows
```{r}
datatable(status[1:100,] %>%                      
    arrange(statusId), 
    options = list(pageLength = 10, scrollX = T)) 
```

# EDA

## F1 across the World

In the last 70 years, the competition in F1 has been held on 6 continents.
```{r, warning=FALSE, fig.width=10}
#circuits map
#Change longitude column name for cohesion
colnames(results_races)[27] <- "lon"
# generate worldwide dots
lat <- data_frame(lat = seq(-90, 90, by = 1.5))
lon <- data_frame(lon = seq(-170, 180, by = 1.5))
dots <- merge(lat, lon, all = TRUE)  

# exclude water-based dots 
dots %>%
  mutate(country = map.where("world", lon, lat),
         lakes = map.where("lakes", lon, lat)) %>%
  filter(!is.na(country) & is.na(lakes)) %>% 
  select(-lakes) ->
  dots

# plot the data
ggplot(dots, aes(x = lon, y = lat)) + 
geom_point(col = "white", size = 0.25) +
geom_point(data = results_races, colour = "red", alpha = .009, size = 5) + 
theme_void() +
theme(panel.background = element_rect(fill = "#484848"),
      plot.background = element_rect(fill = "#484848"))
```
