---
title: "An interactive Formula 1 Analysis"
author: "Nomthunzi Moyo and Mateusz Majak"
date: "22 11 2020"
output: 
  html_document:
    number_sections: true
    fig_caption: true
    toc: true
    fig_width: 7
    fig_height: 4.5
    theme: cosmo
    highlight: tango
    code_folding: hide
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# About the dataset
The dataset used in this project consists of information about F1 races statistics
from 1950 until the latest 2020 season. It was shared by Vopani on the Kaggle 
platform and is available [here](https://www.kaggle.com/rohanrao/formula-1-world-championship-1950-2020).
The whole data comes from [ergast website](http://ergast.com/mrd/) and is divided into
13 csv files:

* circuits
* constructor_results
* constructor_standings
* constructors
* driver_standings
* drivers
* lap_times
* pit_stops
* qualifying
* races
* results
* seasons
* status

With such amount of data, we will try to answer the following questions:

* **How has the competition and level of F1 changed over the years?**
* **Is F1 currently being won in qualifying?**
* **Do the differences between the cars decrease over the years?**
* **Which F1 circuits provide a more even fight?**

# Preparations {.tabset .tabset-fade .tabset-pills}

## Load libraries
We load multiple libraries for data manipulation and visualisation together with 
tools for data modelling.

```{r message = FALSE}

# data manipulation
library(dplyr)
library(tidyr)

# Interactivity
library(plotly)
library(crosstalk)
library(DT)

#Shiny App
library(shiny)

#visualisations
library(ggplot2)
library(RColorBrewer)
library(ggthemes)
```

## Load Data
```{r message = FALSE}

```

# Dataset preparation

# EDA

